---
title: "Week 5 | Tidy Data"
author: "Chaitanya Kundety"
date: "June 2, 2019"
output: html_document
book: "YaRr! The pirate's guide to R"
book-url: "https://bookdown.org/ndphillips/YaRrr/test-your-r-might-pirates-and-superheroes.html
"
---

## 8.7 Exercise

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
```

### 1. Combine the data into a single dataframe. Complete all the following exercises from the dataframe!

```{r}
superhero <- data.frame(
  Name = c("Astrid", "Lea", "Sarina", "Remon", "Letizia", "Babice", "Jonas", "Wendy", "Niveditha", "Gioia"),
  Sex = c("F", "F", "F", "M", "F", "F", "M", "F", "F", "F"),
  Age = c(30, 25, 25, 29, 22, 22, 35, 19, 32, 21),
  Superhero = c("Batman", "Superman", "Batman", "Spiderman", "Batman",
               "Antman", "Batman", "Superman", "Maggott", "Superman"),
  Tattoos = c(11, 15, 12, 5, 65, 3, 9, 13, 900, 0)
)

kable(superhero)
```

### 2. What is the median age of the 10 pirates?

```{r}
med_age <- median(superhero$Age)
```

The median age is `r med_age`.

### 3. What was the mean age of female and male pirates separately?

```{r}
mean_ages <- superhero %>% group_by(Sex) %>% summarize(mean_age = mean(Age))
kable(mean_ages)
```

### 4. What was the most number of tattoos owned by a male pirate?

```{r}
max_tattoos_m <- superhero %>%
  filter(Sex == "M") %>%
  summarize(max(Tattoos))
```

The highest number of tattoos among males is `r max_tattoos_m`.

### 5. What percent of pirates under the age of 32 were female?

```{r}
age_lt32_f_pct <- superhero %>%
  filter(Age < 32) %>%
  group_by(Sex) %>%
  summarize(cnt = n()) %>%
  mutate(pct = cnt / sum(cnt)) %>%
  filter(Sex == "F") %>%
  select(pct)
```

`r age_lt32_f_pct * 100`% of the pirates under age 32 are female.

### 6. What percent of female pirates are under the age of 32?

```{r}
f_lt32_pct <- superhero %>%
  filter(Sex == "F") %>%
  mutate(under32 = Age < 32) %>%
  group_by(under32) %>%
  summarise(cnt = n()) %>%
  mutate(pct = cnt / sum(cnt)) %>%
  filter(under32 == TRUE) %>%
  select(pct)
```

`r f_lt32_pct * 100`% of females are under 32 years of age.

### 7. Add a new column to the dataframe called tattoos.per.year which shows how many tattoos each pirate has for each year in their life.

```{r}
superhero <- superhero %>% mutate(tattoos.per.year = Tattoos / Age)
kable(superhero)
```

### 8. Which pirate had the most number of tattoos per year?

```{r}
max_tat_yr <- superhero %>%
  filter(tattoos.per.year == max(tattoos.per.year)) %>%
  select(Name)
```

`r max_tat_yr` has the most tattoos per year.

### 9. What are the names of the female pirates whose favorite superhero is Superman?

```{r}
f_sups <- superhero %>%
  filter(Sex == "F", Superhero == "Superman") %>%
  pull(Name)
```

`r as.character(f_sups)` are famale wih Superman for superhero.

### 10. What was the median number of tattoos of pirates over the age of 20 whose favorite superhero is Spiderman?

```{r}
age_gt20_spdrmn_med_tat <- superhero %>%
  filter(Age > 20, Superhero == "Spiderman") %>%
  summarize(median(Tattoos))
```

Median number of tattoos of pirates over age 20 and like Spiderman is `r age_gt20_spdrmn_med_tat`.
